openTSNE: a modular Python library for t-SNE dimensionality reduction and embedding

### 2020-07-06 paper/policar2020-opentsne.tex
Improve titles

\subsection*{Uncovering Structure in High-Dimensional Data Sets}

Dimensionality reduction techniques implicitly assume that high-dimensional data lies on a lower-dimensional manifold, which can accurately be captured by a smaller number of dimensions. However, there is no evidence to suggest that any data set can accurately be described using only two dimensions and that any projection onto a two-dimensional plane will inevitably lead to loss of information. Therefore, it is beneficial to examine multiple embeddings, each of which provides a different perspective into the topology and other characteristics of the data. For instance, in Fig.~\ref{fig:tasic}, we show four embeddings of the same gene-expression data from the mouse brain~\cite{tasic2018shared}.
\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{tasic2018}
  \caption{\label{fig:tasic}We use openTSNE to create four different visualizations of the Tasic \textit{et al.}~\cite{tasic2018shared} data set containing 21,874 single-cells originating from the mouse neocortex. Cluster annotations and colors are taken from the original publication. Warm colors correspond to excitatory neurons, cool colors correspond to inhibitory neurons, and gray/brown colors correspond to non-neuronal cells. The standard t-SNE \textbf{(a)} emphasizes local structure while increasing perplexity \textbf{(b)} results in a more meaningful layout of the clusters. We can also combine the two perplexities by using a multiscale kernel affinity model \textbf{(c)} and obtain a good trade-off between global and local structure. Alternatively, we can inspect more fine-grained structure and reveal smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
\end{figure*}
Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters, which uncovers numerous clusters. Warm colors correspond to excitatory neurons, while cool colors represent inhibitory neurons. While different clusters of excitatory and inhibitory neurons appear close to one another, all clusters appear somewhat equidistant from their neighboring clusters, and the overall relations between them are not obvious. On the other hand, Fig.~\ref{fig:tasic}.b focuses on preserving larger neighborhoods of points, resulting in a more globally consistent layout where relations between clusters become more apparent. In this embedding, it is evident from the increased white-space between groups that there is one large class of excitatory neurons, and two related classes of inhibitory neurons. Unfortunately, focusing on preserving large neighborhoods often results in smaller clusters being absorbed into larger ones. Alternatively, Fig.~\ref{fig:tasic}.c uses multi-scale similarity kernels which aim to preserve both the global organization of clusters as well as prevent smaller clusters from being absorbed into larger ones. On the other hand, Fig.~\ref{fig:tasic}.d shows use the same parameter values as Fig.~\ref{fig:tasic}.a, but at a finer level of resolution. This reveals that many of the clusters are composed of other subgroups.

@@

This section is meant to familiarize the reader with the t-SNE algorithm and its recent extensions. This brief review is by no means comprehensive, but is self contained and introduces the mathematics upon which the \opentsne\ library is based.

\subsection*{Preliminaries}
This section is meant to familiarize the reader with the t-SNE algorithm and its recent extensions. This brief review is by no means comprehensive, but is self-contained and introduces the mathematics upon which the \opentsne\ library is based.

@@

\subsection*{Efficient Implementation}
A direct evaluation of t-SNE gradients requires $\mathcal{O}(N^2)$ operations, which makes its application impractical to any reasonably-sized data set and beckons for the development of efficient approximation schemes. Van der Maaten observed that the t-SNE gradient may be reformulated in terms of a N-body problem where data points may be interpreted as particles which attract and repel each other~\cite{van2014accelerating}. The gradient may be rewritten as
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \left [ \sum_{j \neq i} p_{ij} q_{ij} Z \left ( \mathbf{y}_i - \mathbf{y}_j \right ) -\sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right ) \right ], \label{eq:grad_attr_rep}
\end{equation}
where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right )^{-1}$. The two terms may be viewed as the attractive and repulsive forces between particles, respectively. Each term lends itself to efficient approximations, which enable us to greatly reduce the time complexity of t-SNE.

@@

Standard t-SNE is often used to reveal the clustering structure in data at one level of resolution. Different perplexity parameter values can be used to identify global cluster relationships or small, well-isolated groups. Unfortunately, varying perplexity values can be time-consuming as this involves recomputing the KNNG which is often the most expensive part of the t-SNE algorithm. Alternatively,  Kobak \textit{et al.}~\cite{kobak2019heavy} suggest that varying the degree of freedom in the t-distribution used in the embedding space during optimization can be used to explore the clustering structure at different levels of resolution.

@@

\subsection*{Alternative Initialization Schemes}

@@

\begin{table}

@@

\begin{tabular}{r c c c c c c}
\setlength\tabcolsep{6pt}
& \rot{scikit-learn} & \rot{MulticoreTSNE} & \rot{FIt-SNE} & \rot{openTSNE} \\
\toprule
PyPI package & \checkmark & \checkmark & & \checkmark \\
conda package & \checkmark & & & \checkmark \\
Precompiled binaries & \checkmark & & & \checkmark \\
\hline
$\mathcal{O}(N \log N)$ & \checkmark & \checkmark & & \checkmark \\
$\mathcal{O}(N)$ & & & \checkmark & \checkmark \\
\hline
Variable degrees of freedom & & & \checkmark & \checkmark \\
Variable exaggeration & & & \checkmark & \checkmark \\
Multiscale Gaussian kernels & & & \checkmark & \checkmark \\
Custom affinity kernels & & & & \checkmark \\
Adding new samples & & & & \checkmark \\
\bottomrule
\end{tabular}
\end{table}

@@

% or include bibliography directly:
% \begin{thebibliography}
% \bibitem{b1}
% \end{thebibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Figures                       %%
%%                               %%
%% NB: this is for captions and  %%
%% Titles. All graphics must be  %%
%% submitted separately and NOT  %%
%% included in the Tex document  %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%
%% Do not use \listoffigures as most will included as separate files

\section*{Figures}
  \begin{figure}[h!]
  \caption{\csentence{Sample figure title.}
      A short description of the figure content
      should go here.}
      \end{figure}

\begin{figure}[h!]
  \caption{\csentence{Sample figure title.}
      Figure legend text.}
      \end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Tables                        %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
\begin{table}[h!]
\caption{Sample table title. This is where the description of the table should go.}
      \begin{tabular}{cccc}
        \hline
           & B1  &B2   & B3\\ \hline
        A1 & 0.1 & 0.2 & 0.3\\
        A2 & ... & ..  & .\\
        A3 & ..  & .   & .\\ \hline
      \end{tabular}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Additional Files              %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Additional Files}
  \subsection*{Additional file 1 --- Sample additional file title}
    Additional file descriptions text (including details of how to
    view the file, if it is in a non-standard format or the file extension).  This might
    refer to a multi-page table or a figure.

  \subsection*{Additional file 2 --- Sample additional file title}
    Additional file descriptions text.

>>>>

\subsection*{Uncovering Structure in High-Dimensional Data}

Dimensionality reduction techniques implicitly assume that high-dimensional data lies on a lower-dimensional manifold, which can accurately be captured by a smaller number of dimensions. However, there is no evidence to suggest that any data set can accurately be described using only two dimensions and that any projection onto a two-dimensional plane will inevitably lead to loss of information. Therefore, it is beneficial to examine multiple embeddings, each of which provides a different perspective into the topology and other characteristics of the data.
\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{tasic2018}
  \caption{\label{fig:tasic}We use openTSNE to create four different visualizations of the Tasic \textit{et al.}~\cite{tasic2018shared} data set containing 21,874 single-cells originating from the mouse neocortex. Cluster annotations and colors are taken from the original publication. Warm colors correspond to excitatory neurons, cool colors correspond to inhibitory neurons, and gray/brown colors correspond to non-neuronal cells. The standard t-SNE \textbf{(a)} emphasizes local structure while increasing perplexity \textbf{(b)} results in a more meaningful layout of the clusters. We can also combine the two perplexities by using a multiscale kernel affinity model \textbf{(c)} and obtain a good trade-off between global and local structure. Alternatively, we can inspect more fine-grained structure and reveal smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
\end{figure*}
For instance, in Fig.~\ref{fig:tasic}, we show four embeddings of the same gene-expression data from the mouse brain~\cite{tasic2018shared}. Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters, which uncovers numerous clusters. Warm colors correspond to excitatory neurons, while cool colors represent inhibitory neurons. While different clusters of excitatory and inhibitory neurons appear close to one another, all clusters appear somewhat equidistant from their neighboring clusters, and the overall relations between them are not obvious. On the other hand, Fig.~\ref{fig:tasic}.b focuses on preserving larger neighborhoods of points, resulting in a more globally consistent layout where relations between clusters become more apparent. In this embedding, it is evident from the increased white-space between groups that there is one large class of excitatory neurons, and two related classes of inhibitory neurons. Unfortunately, focusing on preserving large neighborhoods often results in smaller clusters being absorbed into larger ones. Alternatively, Fig.~\ref{fig:tasic}.c uses multi-scale similarity kernels which aim to preserve both the global organization of clusters as well as prevent smaller clusters from being absorbed into larger ones. On the other hand, Fig.~\ref{fig:tasic}.d shows use the same parameter values as Fig.~\ref{fig:tasic}.a, but at a finer level of resolution. This reveals that many of the clusters are composed of other subgroups.

@@

This section is meant to familiarize the reader with the t-SNE algorithm and its recent extensions. This brief review is by no means comprehensive, but is self-contained and introduces the mathematics upon which the \opentsne\ library is based.

@@

\subsection*{Efficient Approximation Schemes}
A direct evaluation of t-SNE gradients requires $\mathcal{O}(N^2)$ operations, which makes its application impractical to any reasonably-sized data set and beckons for the development of efficient approximation schemes. Van der Maaten observed that the t-SNE gradient may be reformulated in terms of a N-body problem where data points may be interpreted as particles which attract and repel each other~\cite{van2014accelerating}. The gradient may be rewritten as
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \left [ \sum_{j \neq i} p_{ij} q_{ij} Z \left ( \mathbf{y}_i - \mathbf{y}_j \right ) -\sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right ) \right ], \label{eq:grad_attr_rep}
\end{equation}
where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right )^{-1}$. The two terms may be viewed as the attractive and repulsive forces between particles, respectively. Each term lends itself to efficient approximations, which enable us to greatly reduce the time complexity of t-SNE.

@@

Standard t-SNE reveals the clustering structure in data at one level of resolution. Different perplexity parameter values can be used to identify global cluster relationships or small, well-isolated groups. Unfortunately, varying perplexity values can be time-consuming as this involves recomputing the KNNG which is often the most expensive part of the t-SNE algorithm. Alternatively, Kobak \textit{et al.}~\cite{kobak2019heavy} suggest that varying the degree of freedom in the t-distribution used in the embedding space during optimization can be used to explore the clustering structure at different levels of resolution.

@@

\subsection*{Globally Consistent Initialization Schemes}

@@

\begin{table}
\caption{\label{tab:features}Feature list...}

@@

\begin{tabular}{r c c c c c c}
\toprule
\setlength\tabcolsep{6pt}
& \rot{\textsf{scikit-learn}} & \rot{\textsf{MulticoreTSNE}} & \rot{\textsf{FIt-SNE}} & \rot{\textsf{openTSNE}} \\
\toprule
PyPI package & \checkmark & \checkmark & & \checkmark \\
conda package & \checkmark & & & \checkmark \\
Precompiled binaries & \checkmark & & & \checkmark \\
\hline
Barnes-Hut ($\mathcal{O}(N \log N)$) & \checkmark & \checkmark & & \checkmark \\
FIt-SNE ($\mathcal{O}(N)$) & & & \checkmark & \checkmark \\
\hline
Multiscale Gaussian kernels & & & \checkmark & \checkmark \\
Fully-custom affinity kernels & & & & \checkmark \\
Variable degrees of freedom & & & \checkmark & \checkmark \\
Variable exaggeration & & & \checkmark & \checkmark \\
Better initialization & & & \checkmark & \checkmark \\
Automatic learning rate & & & \checkmark & \checkmark \\
Embedding new samples & & & & \checkmark \\
\bottomrule
\end{tabular}
\end{table}

@@
